\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-31 metatheory-condensed.tex
stuff

W-types~\cite{martin1984intuitionistic} are a standard way to model
inductive types in MLTT.
A constructor of an inductive type~$T$ is given by a pair of types
$\goodType{\Gm}{A}{}$ and $\goodType{\Gm,A}{B}{}$ (rule \ruleref{wsig/add}):
given two arguments $\goodTerm{\Gm}{t_1}{A}$ and
$\goodTerm{\Gm,\sub{B}{\SubstExt{\SubstWeak{0}}{t_1}}}{t_2}{T}$,
it constructs a term of type~$T$ (rule \ruleref{tm/wsup}).
As the rules suggest, our formulation additionally supports
\emph{W-type signatures}.
A well formed W-type signature \goodWSig{\Gm}{\wsig}{n} is
composed of $n$ pairs of types, each of which represents a constructor
(rules \ruleref{wsig/empty} and \ruleref{wsig/add}).

@@

>>>>

W-types~\cite{martin1984intuitionistic} are a succinct way to model
inductive types in MLTT;
together with the identity type $\TyId{\cdot}{\cdot}$, they can express a
whole host of inductive types~\cite{hugunin2021not}.
%
A constructor is given by a pair of types
$\goodType{\Gm}{A}{}$ and $\goodType{\Gm,A}{B}{}$ (rule \ruleref{wsig/add}):
given two arguments $\goodTerm{\Gm}{t_1}{A}$ and
$\goodTerm{\Gm,\sub{B}{\SubstExt{\SubstWeak{0}}{t_1}}}{t_2}{T}$,
$\Wsup{\wsig}{t_1}{t_2}$ constructs a term of type~$T$ (rule \ruleref{tm/wsup}).
As the rules suggest, our formulation additionally supports
\emph{W-type signatures}.
A well formed W-type signature \goodWSig{\Gm}{\wsig}{n} is
composed of $n$ pairs of types, each of which represents a constructor
(rules \ruleref{wsig/empty} and \ruleref{wsig/add}),
with $\wsigproj{i}{1}{\wsig}$ and $\wsigproj{i}{2}{\wsig}$ giving the
$i$-th pair in~$\wsig$, and $\wcode{\wsig}$ giving the code of the W-type (\ruleref{tm/w}).

@@

Well-formedness judgments of \emph{linkage signatures} have form
$\goodSig{\Gm}{\lsig}{n}$.
As rules \ruleref{lsig/empty} and \ruleref{lsig/add} show, a linkage
signature is a list of $n$~types.
%
Well-formedness judgments of \emph{linkages} have form $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$,
where $\TyLkg{\lsig}$ is a type formed by~$\lsig$.
As rules \ruleref{l/empty} and \ruleref{l/add} show, a linkage is a list
of $n$~terms, each representing a field of the family modeled by the linkage.

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the fields preceding the current field~$t$
(i.e., those recorded by the first premise $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$).
Crucially, unlike the second field in a dependent pair (\ruleref{tm/pair}),
the typing of~$t$ in \ruleref{l/add} is polymorphic to how the fields in
$t$'s context are defined, so that a different linkage that overrides
fields in $t$'s context can reuse~$t$---without retyping it---by
projecting it (\ruleref{l/proj2}) and then including it.

This type~$A$ abstracting the context in \ruleref{l/add} and
\ruleref{lsig/add} is not necessarily the same as those recorded by~$\lsig$,
because a field representing the code $\wcode{\wsig}$ of a W-type has to
expose different types to different succeeding fields.
Succeeding fields that invoke
$\Wsup{\wsig}{\cdot}{\cdot}$ or $\Wrec{\wsig}{\cdot}{\cdot}$
should see the concrete signature~$\wsig$, as rules \ruleref{tm/wsup}
and \ruleref{tm/wrec} stipulate, whereas $\wsig$ should be hidden
from all other fields so that they can be reused
for a W-type signature~$\wsig'$ extending~$\wsig$.

We illustrate this with \cref{fig:stlc-linkage-typing}.
On the left is code excerpted from \cref{fig:stlc-mechanized}, and on the right
is how the corresponding fields are modeled and typed in \TT.
%
$\lkg_i$ is the linkage that adds the $i$-th field~$t_i$ with the typing
$\goodTerm{A_i}{t_i}{T_i}$.
%
Here, we feel free to use a notation with explicit names.
We name the context as $\mathit{self}$ (cf.~\cref{fig:stlc-compiled})
and feel free to write its type, which is a dependent tuple type, as a
dependent record type with explicit labels.
%
\begin{itemize}[labelsep=*,leftmargin=1pc]

\item 
The first field $t_1$ is defined as $\wcode{\wsig_{\texttt{tm}}}$
and given the singleton type $\TyS{\wcode{\wsig_{\texttt{tm}}}}$, where $\wsig_{\texttt{tm}}$
is the W-type signature for \lsti{tm}.
This typing is recorded by $\lsig_1$ and is thus available in all $\lsig_i$'s.

\item 
The next four fields model the four constructors of~\lsti{tm}.
Constructor \lsti{tm_unit} is modeled as $\Wsup{\wsig_{\texttt{tm}}}{\top}{\bot}$
and typed with $\El{\mathit{self}.\texttt{tm}}$, where $\mathit{self}$ stands for
the typing context containing the first field~\lsti{tm}.
Because the W-type signature $\wsig_{\texttt{tm}}$ is exposed to this typing context,
$\El{\mathit{self}.\texttt{tm}}$ and $\El{\wcode{\wsig_{\texttt{tm}}}}$ can be equated,
as required by rule \ruleref{tm/wsup}.

\item
Similarly, $\wsig_{\texttt{tm}}$ is exposed to the typing context of
$t_{10}$, which models the recursive function \lsti{subst} by invoking
the recursor $\Wrec{\wsig_{\texttt{tm}}}{\cdot}{\cdot}$.
%
Other fields automatically generated by the plugin, such as partial
recursors and various propositional equalities, can also be modeled in
this way by invoking $\texttt{Wrec}$.

\item
By contrast, $\wsig_{\texttt{tm}}$ is hidden from the typing context of all other fields.
Their typing depends on the knowledge that \lsti!tm! has type~$\cU$,
rather than \TyS{\wcode{\wsig_{\texttt{tm}}}},
so they can be reused in a context where \lsti!tm! is defined as $\wcode{\wsig'_{\texttt{tm}}}$,
where $\wsig'_{\texttt{tm}}$ extends~$\wsig_{\texttt{tm}}$ with additional constructors.

\end{itemize}

The third premise
$\goodTerm{\Gm,\TyTkg{\lsig}}{\seal}{\sub{A}{\SubstWeak{1}}}$
in rules \ruleref{l/add} and \ruleref{lsig/add} is
responsible for hiding W-type signatures in~$\lsig$.
Here, $\TyTkg{\lsig}$ is a dependent tuple type (rule \ruleref{tyeq/pk/add})
packaging the types of all fields preceding the current field~$t$,
and $\seal$ is the term turning the tuple $\Tkg{\lkg}$ of type
$\TyTkg{\lsig}$ (rule \ruleref{tmeq/pk/add}) into a new term of type $A$
that hides W-type signatures, if necessary.
When no hiding is needed, $\seal$ is the identity $\var{0}$ and $A$ is
$\TyTkg{\lsig}$.

\bigskip

A supplemental appendix gives a more complete example showing how families
are modeled and type-checked as linkages.

\begin{figure}
\small
\newcolumntype{L}{>{$}l<{$}}
\newcolumntype{C}{>{$}c<{$}}
\newcolumntype{R}{>{$}r<{$}}
\begin{tabular}{@{}lR@{\ }C@{\ }LR@{\ \,}C@{\,\ }L@{}}
\lsti!Family STLC.!
& \lkg_0 & \coloneqq & \LkgEmp
\\
\lsti!FInductive tm :=!
& \lkg_1 & \coloneqq & \LkgAdd{\lkg_0}{\dadada}
& \mathit{self} : []
& \vdash
& \wcode{\wsig_{\texttt{tm}}} : \TyS{\wcode{\wsig_{\texttt{tm}}}}
\\
\lsti!| tm_unit : tm | !\dadada
& \lkg_2 & \coloneqq & \LkgAdd{\lkg_1}{\dadada}
& \mathit{self} : [\texttt{tm}:\TyS{\wcode{\wsig_{\texttt{tm}}}}]
& \vdash
& \Wsup{\wsig_{\texttt{tm}}}{\top}{\bot} : \El{\mathit{self.}\texttt{tm}}
\\
\lsti!FRecursion subst !\dadada
\\
\lsti!Case tm_unit := !\dadada
& & &
& \mathit{self} : [\texttt{tm}:\cU, \texttt{tm\_unit}: \El{\texttt{tm}}, \dadada]
& \vdash
& \dadada : \CaseSig{\top}{\bot}{\texttt{id}\to\El{\textit{self.}\texttt{tm}}\to\El{\textit{self.}\texttt{tm}}}
\\
\lsti!End subst.!
& & &
& \mathit{self} : [\texttt{tm}:\TyS{\wcode{\wsig_{\texttt{tm}}}}, \texttt{tm\_unit}: \El{\texttt{tm}}, \dadada]
\\
\lsti!!
\end{tabular}
\caption{Typing of each field}
\label{fig:stlc-linkage-typing}
\end{figure}

### 2022-10-31 stlc-mechanized.tex
stlc compilation figure
simplify
stuff

FInductive hasty : env -> tm -> ty -> Prop :=
| ht_var  : [@\dadada@][@\hfill\codecomment{Typing rules}@]
| ht_unit : \forall G, hasty G tm_unit ty_unit
| ht_abs  : [@\dadada@]
| ht_app  : [@\dadada@].

@@

\colorlet{codecomment-color}{black}

@@

\makeline[.5\textwidth+9pt]{Check-STLCFix-typesafe}{Check-STLCFix-typesafe}[codecomment-color!35]

>>>>

FInductive hasty : env -> tm -> ty -> Prop :=
| ht_unit : \forall G, hasty G tm_unit ty_unit
| ht_var  : [@\dadada@][@\hfill\codecomment{Typing rules}@]
| ht_abs  : [@\dadada@]
| ht_app  : [@\dadada@].

@@

\definecolor{codecomment-color}{HTML}{000080}

@@

\makeline[.5\textwidth+9pt]{Check-STLCFix-typesafe}{Check-STLCFix-typesafe}[codecomment-color!50]
